---
title: "Group Project"
format:
 pdf:
  fontsize: 10pt
   
editor: visual
---

## Introduction

Obesity is a major public health concern in Scotland and is regularly monitored through the Scottish Health Survey. Examining changes in obesity prevalence over time can help identify trends, and determine whether the situation is improving, worsening, or remaining stable. This analysis investigates whether the prevalence of obesity in Scotland changed between 2013 and 2016, and also examines differences in obesity prevalence across age group, sex, employment status, and lifestyle factors such as fruit and vegetable intake.

Questions of Interest:

1.  Has the prevalence of obesity in Scotland changed over the given years of the Scottish Health Survey?

2.  Are there any differences in obesity by age, gender, socio-economic status or lifestyle factors?

## Exploratory Analysis

```{r}
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(ggplot2)
library(broom)
library(yardstick)  

health<- read.csv("DAProject3.csv")
```

The relative proportions of the response variable, obesity status (Yes/No), are compared across years to identify any noticeable changes in obesity prevalence over time. Overall, approximately 30% of individuals in the dataset are classified as obese. The plots below illustrate how obesity prevalence varies across the groups considered.

```{r}
#| echo: false

health.obesity<- health %>%
  select(Obese, Year)
```


### Prevalence of Obesity in Scotland

```{r}
#| echo: false
#| label: barplot1
#| fig-cap: "Barplot of Obesity Prevalence for the Years (2013–2016)"
#| fig-width: 5
#| fig-height: 1.5
ggplot(health.obesity, aes(x = Year, group = Obese)) +
    geom_bar(aes(y = after_stat(prop), fill = Obese), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Obese", title = "Obesity Prevalence for the Years(2013-2016)")

```

The bar plot shows a consistent pattern across the years 2013–2016. In each year, the distribution between the two categories remains very similar. The proportion classified as obese (“Yes”) is approximately 69–70%, while those classified as not obese (“No”) account for about 30–31%. Overall, the plot indicates that obesity prevalence remains relatively stable across the years shown.


```{r}
#| echo: FALSE
#| output: FALSE
library(tidyverse)
library(ggplot2)

data <- read.csv("DAProject3.csv", stringsAsFactors = TRUE)
```

```{r}
#| echo: FALSE
data$Obese <- factor(data$Obese)
```

### Obesity vs Sex

```{r}
#| echo: false
#| fig-cap: "Obesity vs Sex"
#| fig-width: 2.5
#| fig-height: 1.5
#| fig-align: center
data$Sex <- factor(data$Sex)
ggplot(data, aes(x = Sex, fill = Obese)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Sex",
    y = "Percentage"
  )

```

The percentage of individuals classified as obese is similar for males and females. Around 31% of females are classified as obese compared with about 29% of males. This suggests that the difference in obesity prevalence between the two groups is relatively small.

### Obesity vs Age Group

```{r}
#| echo: false
#| fig-cap: "Obesity vs Age Group"
#| fig-width: 5
#| fig-height: 1.5
#| fig-align: center
data$AgeGroup <- factor(data$AgeGroup)
ggplot(data, aes(x = AgeGroup, fill = Obese)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Age Group",
    y = "Percentage"
  )
```

Clear differences in obesity prevalence can be observed across age groups. Approximately 16% of individuals aged 16–24 are classified as obese, and this increases steadily across older age groups. The highest prevalence appears in the 65–74 age group at around 36–37%. The prevalence then decreases slightly to about 29% in the 75+ group.

### Obesity vs Employment

```{r}
#| echo: false
#| fig-cap: "Obesity vs Employment"
#| fig-width: 6
#| fig-height: 2.5
#| fig-align: center
data$Employment <- factor(data$Employment)
ggplot(data, aes(x = Employment, fill = Obese)) +
  geom_bar(position = "fill") +
  labs(
    title = "Obesity vs Employment",
    x = "Employment",
    y = "Percentage"
  ) +
  scale_y_continuous(labels = scales::percent) +
  coord_flip()
```

Differences in obesity prevalence are visible across employment categories. Individuals permanently unable to work show the highest obesity prevalence at roughly 42–43%. Those looking after home or family also have relatively high obesity levels at around 37%. In contrast, individuals in full-time education have the lowest prevalence at approximately 17%.

### Obesity vs Lifestyle Choices


```{r}
#| echo: false
#| fig-cap: "Obesity vs Intake"
#| fig-height: 1.5
#| fig-width: 2.5
#| layout-ncol: 2
ggplot(data, aes(x = Veg, fill = Obese)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent)

ggplot(data, aes(x = Fruit, fill = Obese)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent)
```

Lifestyle factors show only small differences in obesity prevalence. Individuals who meet the recommended fruit intake and those who do not both have obesity levels of around 30%. A similar pattern is observed for vegetable intake, where those meeting the recommendation have obesity prevalence of about 30%, compared with roughly 33% among those who do not. Overall, the differences between the groups are relatively small.

