---
title: "Group Project"
format: html
editor: visual
---

# Introduction

Obesity is a major public health concern in Scotland and is regularly monitored through the Scottish Health Survey. Examining changes in obesity prevalence over time can help identify trends, and understand whether the situation is improving, worsening, or staying the same. This analysis explores whether the prevalence of obesity in Scotland changed between 2013 and 2016. The proportions for each year are compared using the relative percentages of the \"Yes\" factor (individuals classified as obese) and the \"No\" factor (individuals not classified as obese) to determine whether there were any noticeable changes in prevalence across the given years.

***Question of interest:***

***Has the prevalence of obesity in Scotland changed over the given years of the Scottish Health Survey?***

# Exploratory Analysis

```{r}
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(ggplot2)
library(sjPlot)
library(broom)
library(performance)
library(yardstick)  

health<- read.csv("DAProject3.csv")
```

```{r}
#| echo: false

health.obesity<- health %>%
  select(Obese, Year)
```

```{r}
#| echo: false
#| message: false
#| warning: false
library(janitor)
health.obesity %>%
  tabyl(Year, Obese) %>%
  adorn_percentages() %>%
  adorn_pct_formatting() %>%
  adorn_ns() # To show original counts
```

This output above shows the percentage data in an organised way.

```{r}
#| echo: false
#| label: barplot1
ggplot(health.obesity, aes(x = Year, group = Obese)) +
    geom_bar(aes(y = after_stat(prop), fill = Obese), 
             stat = "count", position = "dodge") +
    labs(y = "Proportion", fill = "Obese", title = "Barplot of Obesity for the Years(2013-2016)")

```

From the barplot, we can see that:

1.  In 2013, the proportion of individuals classified as obese was higher for the **\"Yes\" factor**, accounting for **69.7%**, compared with the **\"No\" factor** at **30.3%**.
2.  Similarly, in 2014 the **\"Yes\" factor** continued to represent the larger proportion of obesity at **69.8%**, while the **\"No\" factor** accounted for **30.2%**.
3.  A similar pattern is observed in the following years. In **2015**, the **\"Yes\" factor** represented **69.4%** of the proportion, compared with **30.6%** for the **\"No\" factor**.
4.  In **2016**, the **\"Yes\" factor** again had the higher proportion at **69.5%**, while the **\"No\" factor** accounted for **30.5%**.

Now we shall fit a logistic regression model to determine whether the gender of a teaching instructor can be predicted from their ethnicity.

# Formal Analysis

# Conclusion

Overall, the results suggest that the prevalence of obesity in Scotland remained fairly stable between 2013 and 2016. In each year, the proportion classified as obese was consistently close to 70% with (69.7% in 2013, 69.8% in 2014, 69.4% in 2015, and 69.5% in 2016), while around 30% were not classified as obese. Although there are small differences between the years, these changes are very minor. This suggests that obesity prevalence did not change significantly during this period and remained relatively consistent across the years analysed.
